### 2025-02-18 paper/body_ecml.tex
vamos

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta||_\infty < \varepsilon\}
\end{aligned}
}\end{equation}

Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
an adversarial example iff it is imperceptible, i.e.~the magnitude of
the perturbation of any individual feature is upper-bounded at
\(\varepsilon\).

@@

\begin{theorem}[Protecting Immutable
Features]\protect\hypertarget{thm-mtblty}{}\label{thm-mtblty}

Let
\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
denote a linear classifier with softmax activation \(\mathcal{S}\)
(i.e.~\emph{multinomial logistic regression}) where
\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
we assume multivariate Gaussian class densities with common diagonal
covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
then protecting an immutable feature from the contrastive divergence
penalty (Equation~\ref{eq-div}) will result in lower classifier
sensitivity to that feature relative to the remaining features, provided
that at least one of those is mutable.

\end{theorem}

It is worth highlighting that Theorem~\ref{thm-mtblty} assumes
independence of features. This raises a valid concern about the effect
of protecting immutable features in the presence of proxy features that
remain unprotected. We discuss this limitation in
Section~\ref{sec-discussion}.

@@

We have simulated this scenario using synthetic data with independent
features and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
outcomes for different training procedures using the same model
architecture each time (a linear classifier). In each case, we show the
linear decision boundary (green) and the training data colored according
to their ground-truth label: orange points belong to the target class,
\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
indicate counterfactuals in the target class generated at test time
using generic gradient descent for a fixed number of iterations.

@@

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{body_ecml_files/mediabag/../paper/figures/poc.pdf}}

}

\caption{\label{fig-poc}Visual illustration of how counterfactual
training improves explainability. See Example~\ref{exm-grad} for
details.}

\end{figure}%

@@

\begin{credits}
\subsubsection{\ackname}

- Aleks
- DelftBlue
- Julia packages: Julia paper itself, PrettyTables, Flux, Plots, AlgebraOfGraphics

\subsubsection{\discintname}

The authors have no competing interests to declare that are relevant to the content of this article.
\end{credits}

>>>>

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation}

Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
magnitude of the perturbation to any individual feature is at most
\(\varepsilon\). This is closely aligned with
\cite{szegedy2013intriguing}, who define an adversarial attack as an
``imperceptible non-random perturbation''. Thus, we choose to work with
a different distinction between CE and AE than
\cite{freiesleben2022intriguing}, who considers misclassification as the
key distinguishing feature of AE. One of the key observations in this
work is that we can leverage counterfactual explanations during training
and get adversarial examples, essentially for free.

@@

\begin{proposition}[Protecting Immutable
Features]\protect\hypertarget{prp-mtblty}{}\label{prp-mtblty}

Let
\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
denote a linear classifier with softmax activation \(\mathcal{S}\)
(i.e.~\emph{multinomial logistic regression}) where
\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
we assume multivariate Gaussian class densities with common diagonal
covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
then protecting an immutable feature from the contrastive divergence
penalty (Equation~\ref{eq-div}) will result in lower classifier
sensitivity to that feature relative to the remaining features, provided
that at least one of those is mutable and discriminative.

\end{proposition}

It is worth highlighting that Proposition~\ref{prp-mtblty} assumes
independence of features. This raises a valid concern about the effect
of protecting immutable features in the presence of proxy features that
remain unprotected. We discuss this limitation in
Section~\ref{sec-discussion}.

@@

We have simulated this scenario using synthetic data with independent
features and Gaussian class-conditional densities in \textbf{?@fig-poc}.
The four panels in \textbf{?@fig-poc} show the outcomes for different
training procedures using the same model architecture each time (a
linear classifier). In each case, we show the linear decision boundary
(green) and the training data colored according to their ground-truth
label: orange points belong to the target class, \(y^+=2\), blue points
belong to the non-target class, \(y^-=1\). Stars indicate
counterfactuals in the target class generated at test time using generic
gradient descent for a fixed number of iterations.

@@

@@

### 2025-02-18 paper/ecml/paper.tex
progress on methodology
slowly slowly
first version of research questions defined
fucking latex
latex is going to kill me one day
vamos

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{mwe}
\usepackage{color}
\usepackage[numbers,compress]{natbib}
\usepackage{longtable}
\usepackage{hyperref}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

@@

@@

\begin{credits}
\subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
used for general acknowledgments, for example: This study was funded
by X (grant number Y).

\subsubsection{\discintname}
It is now necessary to declare any competing interests or to specifically
state that the authors have no competing interests. Please place the
statement with a bold run-in heading in small font size beneath the
(optional) acknowledgments,
for example: The authors have no competing interests to declare that are
relevant to the content of this article. Or: Author A has received research
grants from Company W. Author B has received a speaker honorarium from
Company X and owns stock in Company Y. Author C is a member of committee Z.
\end{credits}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{color}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{svg}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

@@

\let\corollary\relax
\let\endcorollary\relax
\newtheorem{corollary}{Research Question}

@@

\begin{credits}
\subsubsection{\ackname}

- Aleks
- DelftBlue
- Julia packages: Julia paper itself, PrettyTables, Flux, Plots, AlgebraOfGraphics

\subsubsection{\discintname}

The authors have no competing interests to declare that are relevant to the content of this article.
\end{credits}

### 2025-02-18 paper/preprint/paper.tex
vamos

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization. Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on, firstly, identifying
similar input samples with different annotations and, secondly, ensuring
that the gradient of the classifier aligns with the vector between pairs
of counterfactual inputs using the cosine distance as a loss function.
In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation: Wu et
al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{POLYJUICE}, a
general-purpose counterfactual generator for language models. They
demonstrate empirically that augmenting training data through
\emph{POLYJUICE} counterfactuals improves robustness in a number of NLP
tasks. Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023})
introduce Counterfactual Adversarial Training (CAT), which also aims at
improving generalization and robustness of language models.
Specifically, they propose to proceed as follows: firstly, they identify
training samples that are subject to high predictive uncertainty;
secondly, they generate counterfactual explanations for those samples;
and, finally, they fine-tune the given language model on the augmented
dataset that includes the generated counterfactuals.

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
target class \(y^+ \sim p(y)\) and then generating a counterfactual
explanation for that target, similar to how conditional sampling is used
to draw from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}). In particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the target class.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or
adversarial examples
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
leaves some flexibility with respect to the exact choice for
\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
functional form to use, though likely not the only reasonable choice, is
inspired by adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
0 & \text{otherwise.}
\end{cases}
\end{aligned}
}\end{equation}

Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
an adversarial example iff it is imperceptible, i.e.~the magnitude of
the perturbation of any individual feature is upper-bounded at
\(\varepsilon\).

@@

@@

In Section~\ref{sec-constraints} we explain that to avoid penalizing
implausibility that arises due to mutability constraints, we impose a
point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
argue in Section~\ref{sec-constraints} that this approach induces models
to be less sensitive to immutable features and demonstrate this
empirically in Section~\ref{sec-experiments}. Below we derive the
analytical results in Theorem~\ref{thm-mtblty}.

@@

This implies that if we do not protect feature \(d_{\text{immtbl}}\),
the contrastive divergence penalty will push down on
\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
effect
\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
In words, not protecting the immutable feature would have the
undesirable effect of making the classifier more sensitive to this
feature, in that it would be more likely to predict class \(y^-\) as
opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).

By the same rationale, the contrastive divergence penalty can generally
be expected to push up on \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
In words, this has the effect of making the classifier more sensitive to
the mutable feature, in that it would be more likely to predict class
\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, breakable, leftrule=.75mm, coltitle=black, bottomtitle=1mm, toprule=.15mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-warning-color!10!white, arc=.35mm, bottomrule=.15mm, left=2mm, titlerule=0mm, opacityback=0, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, toptitle=1mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, breakable, leftrule=.75mm, coltitle=black, bottomtitle=1mm, toprule=.15mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, bottomrule=.15mm, left=2mm, titlerule=0mm, opacityback=0, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm]

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization. Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on, firstly, identifying
similar input samples with different annotations and, secondly, ensuring
that the gradient of the classifier aligns with the vector between pairs
of counterfactual inputs using the cosine distance as a loss function.

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation: Wu et
al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{Polyjuice}, a
general-purpose counterfactual generator for language models. They
demonstrate empirically that augmenting training data through
\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
tasks. Balashankar et al.
(\citeproc{ref-balashankar2023improving}{2023}) also use
\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
and show that classifier robustness improves up to 20\%. Finally, Luu
and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
Counterfactual Adversarial Training (CAT), which also aims at improving
generalization and robustness of language models. Specifically, they
propose to proceed as follows: firstly, they identify training samples
that are subject to high predictive uncertainty; secondly, they generate
counterfactual explanations for those samples; and, finally, they
fine-tune the given language model on the augmented dataset that
includes the generated counterfactuals.

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the target class. We generate samples
\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
that target over \(T\) iterations using a gradient-based counterfactual
generator. This is similar to how conditional sampling is used to draw
from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}).

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023};
\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023}) or
adversarial examples
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
leaves some flexibility with respect to the exact choice for
\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
functional form to use, though likely not the only reasonable choice, is
inspired by adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation}

Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
magnitude of the perturbation to any individual feature is at most
\(\varepsilon\). This is closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}), who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}), who considers
misclassification as the key distinguishing feature of AE. One of the
key observations in this work is that we can leverage counterfactual
explanations during training and get adversarial examples, essentially
for free.

@@

\bibitem[\citeproctext]{ref-balashankar2023improving}
Balashankar, Ananth, Xuezhi Wang, Yao Qin, Ben Packer, Nithum Thain, Ed
Chi, Jilin Chen, and Alex Beutel. 2023. {``Improving Classifier
Robustness Through Active Generative Counterfactual Data
Augmentation.''} In \emph{Findings of the Association for Computational
Linguistics: EMNLP 2023}, 127--39.

@@

In Section~\ref{sec-constraints} we explain that to avoid penalizing
implausibility that arises due to mutability constraints, we impose a
point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
argue in Section~\ref{sec-constraints} that this approach induces models
to be less sensitive to immutable features and demonstrate this
empirically in Section~\ref{sec-experiments}. Below we derive the
analytical results in Proposition~\ref{prp-mtblty}.

@@

This implies that if we do not protect feature \(d_{\text{immtbl}}\),
the contrastive divergence penalty will decrease
\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
effect
\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
In words, not protecting the immutable feature would have the
undesirable effect of making the classifier more sensitive to this
feature, in that it would be more likely to predict class \(y^-\) as
opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).

By the same rationale, the contrastive divergence penalty can generally
be expected to increase \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
In words, this has the effect of making the classifier more sensitive to
the mutable feature, in that it would be more likely to predict class
\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).

@@

\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, titlerule=0mm, opacityback=0, coltitle=black, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, bottomtitle=1mm, toptitle=1mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, breakable, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, titlerule=0mm, opacityback=0, coltitle=black, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, bottomtitle=1mm, toptitle=1mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, arc=.35mm, breakable, leftrule=.75mm]

### 2025-02-18 paper/sections/method.qmd
vamos

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with respect to the exact choice for $\text{advloss}(\cdot)$ in @eq-obj. An intuitive functional form to use, though likely not the only reasonable choice, is inspired by adversarial training:

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}

Under this choice we treat the counterfactual $\mathbf{x}^\prime$ as an adversarial example iff it is imperceptible, i.e. the magnitude of the perturbation of any individual feature is upper-bounded at $\varepsilon$.

@@

::: {#thm-mtblty}

@@

Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e. *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is mutable.

@@

It is worth highlighting that @thm-mtblty assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxy features that remain unprotected. We discuss this limitation in @sec-discussion.

>>>>

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with respect to the exact choice for $\text{advloss}(\cdot)$ in @eq-obj. An intuitive functional form to use, though likely not the only reasonable choice, is inspired by adversarial training:

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}

Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as adversarial examples as long as the magnitude of the perturbation to any individual feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing, who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing, who considers misclassification as the key distinguishing feature of AE. One of the key observations in this work is that we can leverage counterfactual explanations during training and get adversarial examples, essentially for free.

@@

::: {#prp-mtblty}

@@

Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e. *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is mutable and discriminative.

@@

It is worth highlighting that @prp-mtblty assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxy features that remain unprotected. We discuss this limitation in @sec-discussion.
